Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-04-24 src/tex/ms.tex
[showyourwork] overleaf sync

% Patch \hypertarget; for details, see 
% https://tex.stackexchange.com/a/17138/97669
\makeatletter
\newcommand{\affiliationtarget}[1]{\Hy@raisedlink{\hypertarget{#1}{}}}
\makeatother
\newcommand{\affiliationlink}[1]{\hyperlink{#1}{#1}}

% Chasing rainbows with the Habitable Worlds Observatory
\title{Chasing \rainbows{} with the Habitable Worlds Observatory}
%Scattering phase angle probed by directly imaged planets, or
%Don't ask us about contrast

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan$^{\affiliationlink{1},}$\thanks{Correspondence:  \url{sophia.vaughan@physics.ox.ac.uk}}
    Kimberly Bott$^{\affiliationlink{2},\affiliationlink{28}}$,
    Sarah L. Casewell$^{\affiliationlink{3}}$,
    Nicolas B. Cowan$^{\affiliationlink{4}}$,
    David S. Doelman$^{\affiliationlink{5,19}}$,
    \newauthor 
    Timothy D. Gebhard$^{\affiliationlink{6},\affiliationlink{29}}$,
    Matthew Kenworthy$^{\affiliationlink{5}}$,
    Johan Mazoyer$^{\affiliationlink{7}}$,
    Maxwell A. Millar-Blanchaer$^{\affiliationlink{8}}$,
    \newauthor 
    Olivier Absil$^{16}$,
    Lisa Altinier$^{13}$,
    Pierre Baudoz$^{7}$,
    Ruslan Belikov$^{24}$,
    Alexis Bidot$^{9}$,
    Markus J. Bonse$^{29}$,
    \newauthor 
    Bernhard Brandl$^{5}$,
    Alexis Carlotti$^{9}$,
    Elodie Choquet$^{13}$,
    Niyati Desai$^{17}$,
    Kevin Fogarty$^{24}$,
    Jules Fowler$^{12}$,
    \newauthor
    Yann Gutierrez$^{7,14,22}$,
    Olivier Guyon$^{15,25,26,27}$,
    Sebastiaan Y. Haffert$^{15}$,
    Olivier Herscovici-Schiller$^{14}$, 
    \newauthor
    Adrien Hours$^{9}$,
    Roser Juanola-Parramon$^{20,21}$,
    Elina Kleisioti $^{5}$,
    Lorenzo König$^{16}$,
    Mariya Krasteva$^{23}$, 
    \newauthor
    Iva Laginja$^{7}$,
    Rico Landman$^{5}$,
    Lucie Leboulleux$^{9}$,
    David Mouillet$^{9}$,
    Mamadou N’Diaye$^{10}$,
    Emiel H. Por$^{18}$,
    \newauthor
    Laurent Pueyo$^{18}$,
    Frans Snik$^{5}$,
    Daphne M. Stam$^{30}$,
    Victor Trees$^{31,32}$,
    Dirk van Dam$^{5}$,
    Kyle van Gorkom$^{15}$,
    \newauthor
    Maaike van Kooten$^{11}$ 
    \newauthor \\%
    Affiliations are listed at the end of the paper
    % Timmy: I checked MNRAS and it seems that this is how they handle very long author lists. See, for example:
    % https://academic.oup.com/mnras/article/498/2/2354/5900562
    % I've also added some LaTeX magic to create links for the affiliations; if y'all agree that it's a good idea I can extend it to all authors?
}
\date{Draft version: \today}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\hwo), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. 
%
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
%
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
%
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
%
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
%
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
%
The glint signature at planetary phase angles of $\sim\qtyrange{50}{70}{\degree}$ 
%\sim$110--130$^\circ$ 
would be accessible to the Habitable Worlds Observatory in $\sim\num{14}$ systems, and the rainbow of water clouds at phases of \qtyrange{140}{160}{\degree} would be accessible in $\sim\num{42}$ systems, assuming a \qty{62}{\mas} inner working angle (\IWA{}; $3\,\lambda/D$ for a \qty{6}{\meter} telescope at \qty{600}{\nano\meter}).
%
Increasing the \IWA{} to $2\,\lambda/D$ increases the number of systems with detectable rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA{}.  \end{abstract}

@@

%--------------------------------------------------------------------------------
\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{figures/Fig1IWAPol.png}
    \caption{
        An illustration of phase curves for two homogeneous planets in an edge on system with oceans and an Earth-like atmosphere. 
        The cloudy scenario includes an Earth-like water cloud deck at low altitudes. 
        Below the plot, we are showing the features to which the phase angles and scattering angles are the most sensitive. 
        In reality, as can be seen by the examples above, these features overlap.  
        Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (scattering angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest scattering angles).
    }
    \label{fig:bottplot}
\end{figure*}
%--------------------------------------------------------------------------------

Signatures of habitability, from ocean glint to optical effects from condensates---like the rainbow---should be some of the most distinct signals in an exoplanet's reflected light phase curve, particularly in polarization.
%
However, depending on system architectures, some of these features can be obscured by the inner working angle of a coronagraph.

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star, for which a rocky planet (or moon) may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the stellar flux that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and, for example, geological activity. 
%
Establishing the presence of liquid surface water on an exoplanet would therefore provide important context for the interpretation of possible biosignatures.
%
Providing additional context to the habitability of an exoplanet is the presence of a water cycle, as evidenced by the identification of a rainbow and other optical effects that can be attributed to water clouds.

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off smooth liquid surfaces.
%
In the total flux, the glint signal will be most prominent at small planetary phase (large scattering) angles.
%
The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at \qty{53}{\degree} \citep[i.e., a scattering angle of \qty{127}{\degree}; see, e.g.,][]{2008Icar..195..927W}.
%
This angle therefore varies slightly with wavelength and with sea and atmosphere composition.
%
In general, the glint on a planet will be strongest at red wavelengths where the Rayleigh scattering in the planet's atmosphere and attenuation through the atmosphere are smallest \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also small \citep{2012ApJ...752L...3C}.
%
The strengthened glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
{\color{red}{check the next sentence:}}
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across scattering angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

@@

Rainbows, glories and other optical effects from condensates provide a wealth of information about exoplanets \citep{karalidi2012rainbow, stam2008,2014A&A...566L...1G}.
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to haze rather than a surface \citep{zellner1973polarization}, the phase curve behaviour of Venus showed not only that the clouds were $\sim$75\% H$_2$SO$_4$--water mixture but also constrained the droplet size distribution \citep{hansenhovenier1974}. The glory feature in Venus's phase curve was later used to rule out some solutions for the unknown UV absorber \citep{petrova2018glory}.

In unpolarized light, the scattering by cloud particles can dramatically change the shape of a planet's phase curve compared to that of a 
Lambertian reflecting planet.
%
{\color{red}{check the sentence:}}
While the glory and rainbow are technically visible in any phase curve
in the presence of spherical cloud and/or haze particles, their strongly polarized nature coupled with the suppression of signals from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized light (both in the degree of polarization and in full Stokes measurements) \citep{karalidi2011, stam2008, treesandstam2019}.
%
The rainbow angle depends on the refractive index of the scatterers and thus depends on the composition and the wavelength. 
The wavelength dependence itself also depends on the particles' size distribution, and the size distribution affects the angular shape of the rainbow \citep{karalidi2011}
%
Because the rainbow feature is more narrow than the glint, it would 
allow for a more exact determination of the particle composition.
%
Note that the scattering by clouds that consist of solid particles, 
such as ice crystals, will produce other optical features, such as halo's, particularly at small scattering angles \citep{bailey2007, hansentravis1974, Karalidi2012ice}.
%

@@

%--------------------------------------------------------------------------
\begin{table}
    \centering
    \caption{
        Conversion between $\lambda / D$ and \unit{\mas} for wavelengths $\lambda$ of \qty{600}{\nano\meter} and \qty{1}{\micro\meter}, assuming the telescope's diameter $D$ is $\qty{6}{\meter}$ for the Inner and Outer Working Angles (i.e., \IWA{} and \OWA{}, respectively). 
        %\todo{We should use consistent wavelengths for this table, for the Bott Plot, and for the M\&M cumulative distribution function. Also update to use the only wavelength we choose.}
    }
    \label{tab:IWA_OWA}
    \begin{tabular}{ c c c c } 
    \toprule
    & $\lambda/D$ & \unit{\mas} (at \qty{600}{\nano\meter}) & \unit{\mas} (at \qty{1}{\micro\meter}) \\
    \midrule
    \midrule
    IWA & 1 & 20.63 & 34.38 \\
    IWA & 2 & 41.25 & 68.75 \\
    IWA & 3 & 61.88 & 103.13 \\
    IWA & 4 & 82.51 & 137.51 \\
    \midrule
    OWA & 32 &  660.05 & 1100.08 \\
    OWA & 64 & 1320.09 & 2200.16 \\
    \bottomrule
    \end{tabular}
\end{table}
%--------------------------------------------------------------------------

\hwo{} is envisioned to observe the stars in the NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory.%
\footnote{\hwo{} target list: \url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list comprises $\sim\num{160}$ stars, the majority of which are Sun-like dwarfs: 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude of $R_c = 31$ and a contrast floor of \num{2.5e-11}. 
%with an occurrence rate of rocky planets in the optimistic habitable zone to be $\eta_{\oplus}$ = 0.24 \citep{decadal}. 
The corresponding adopted habitable zone limits are a semi-major axis of \qtyrange{0.95}{1.67}{\au} for a solar twin, planet sizes between \qtyrange{0.8}{1.4}{} Earth radii, and for non-solar stars, scale as square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}. 
%
The authors of the Decadal Survey determined that for an imaging survey of habitable zones with a \qty{6}{\meter} telescope, these are the nearby stars (maximum distance \qty{25}{\parsec}) around which exo-Earths would be the best observable.

@@

In the non-eccentric case (i.e., a circular orbit with the coronagraphic mask centred on the star), the minimum ($\varphi_\mathrm{min}$) and maximum ($\varphi_\mathrm{max}$) accessible scattering phase angles are symmetric about 90 degrees. 
%
We define the angle $\Delta \varphi$ such that: 
\begin{equation}
 \label{eq:Delta_phi}
    \Delta \varphi 
    = \varphi_\mathrm{max} - \varphi_\mathrm{min}
    =  2 \cdot (\varphi_\mathrm{max} - \qty{90}{\degree}) 
    =  2 \cdot (\qty{90}{\degree} - \varphi_\mathrm{min} )  \,.
\end{equation} 
The on-sky projected planet distance, $r_\mathrm{proj}$ (in \unit{\au}) is shown in \cref{fig:scattering-angle} for a non-eccentric case with a 
semi-major orbital axis $a$ (in \unit{\au}), at a distance $d_*$ (in \unit{\parsec}) from the observer. 
%
The scattering phase angle $\varphi$ is given by
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a} \,.
\end{equation}
%
Using small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the angular separation on-sky.
%
We access the minimum scattering angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Using \cref{eq:Delta_phi}, we deduce:
%
\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \varphi}{2}\right) = \frac{\mathrm{IWA \cdot d_*}}{a} \,.
\end{equation}

@@

Most planetary systems will not be viewed as edge-on from Earth.
%
For a system with inclination angle $i$, we can identify the following 2 regimes (see also \cref{fig:orb-grid}):
\begin{enumerate}
    \item The coronagraph does not obscure any part of the sky-projected orbit: 
    the maximum scattering phase angle coverage depends solely on $i$. This 
    holds for orbits that are relatively face-on; 
    \item The coronagraph does obscure parts of the sky-projected orbit: 
    the scattering phase angle coverage depends only on the \IWA{} and 
    the orbital semi-major axis $a$. 
\end{enumerate}

@@

The NASA Exoplanet Exploration Program’s Mission Star List for \hwo{} 
includes for every star an estimated contrast for an Earth-twin in 
the habitable zone.
%
% The cited paper has all the equations that lead to this equation:
The contrast is given by \citep[see, e.g.,][]{2023A&A...671A.165M}
\begin{equation}
    C(\alpha) 
    = F_p(\alpha) / F_* 
    = A_\mathrm{G} \cdot \phi(\alpha) \cdot \left( R_p / a(\alpha) \right)^2 \,,
\end{equation}
where $A_\mathrm{G}$ is the planet's geometric albedo, $R_p$ its radius, $\phi(\alpha)$ is the integral phase function at phase angle $\alpha$
(normalized to \num{1.0} at full phase), and $a$ is the distance between the planet and its host star.
\todo{here, $\alpha$ is used for the phase angle ... and is the 
'integral phase function' the same as the 'scattering phase function'?}
%
The assumptions that went into the contrast estimations for \hwo{}
are that $a$ is constant (the orbits are thus circular), the geometric albedo $A_{\rm G}$ is \num{0.2}, and that the scattering phase function, $\phi(\alpha)$, can be described by a simple Lambertian reflectance phase function, which assumes isotropic reflection by the planetary surface.
%
These calculations give an order-of-magnitude calculation for the 
contrast of Earth-like planets around nearby stars.
%
Therefore, in this work, we will assume that the calculated contrasts 
in the list correspond to the reflected flux at quadrature and normalize the scattering phase functions to this point.
%
We do not aim to accurately calculate the absolute scaling of the contrast, as this depends on many factors and would have required some work for the star list to be redone.
%

>>>>

% Patch \hypertarget; for details, see 
% https://tex.stackexchange.com/a/17138/97669
\makeatletter
\newcommand{\affdest}[1]{\Hy@raisedlink{\hypertarget{#1}{}}$^\mathrm{#1}$}
\makeatother
\newcommand{\afflink}[1]{\hyperlink{#1}{#1}}

% Chasing rainbows with the Habitable Worlds Observatory
\title{Chasing \rainbows{} with the Habitable Worlds Observatory}

%Scattering phase angle probed by directly imaged planets, or
%Don't ask us about contrast

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan\thanks{Correspondence:~\url{sophia.vaughan@physics.ox.ac.uk}}\textsuperscript{,\afflink{1}},
    Kimberly Bott\textsuperscript{\afflink{2},\afflink{3},\afflink{4}},
    Sarah L. Casewell\textsuperscript{\afflink{5}},
    Nicolas B. Cowan\textsuperscript{\afflink{6}},
    David S. Doelman\textsuperscript{\afflink{7},\afflink{8}},
    \newauthor
    Timothy D. Gebhard\textsuperscript{\afflink{9},\afflink{10}},
    Matthew Kenworthy\textsuperscript{\afflink{7}},
    Johan Mazoyer\textsuperscript{\afflink{11}},
    Maxwell A. Millar-Blanchaer\textsuperscript{\afflink{12}},
    \newauthor
    Olivier Absil\textsuperscript{\afflink{13}},
    Lisa Altinier\textsuperscript{\afflink{14}},
    Pierre Baudoz\textsuperscript{\afflink{11}},
    Ruslan Belikov\textsuperscript{\afflink{15}},
    Alexis Bidot\textsuperscript{\afflink{16}},
    Markus J. Bonse\textsuperscript{\afflink{10}},
    \newauthor
    Bernhard Brandl\textsuperscript{\afflink{7}},
    Alexis Carlotti\textsuperscript{\afflink{16}},
    Elodie Choquet\textsuperscript{\afflink{14}},
    Niyati Desai\textsuperscript{\afflink{17}},
    Kevin Fogarty\textsuperscript{\afflink{15}},
    Jules Fowler\textsuperscript{\afflink{18}},
    \newauthor
    Yann Gutierrez\textsuperscript{\afflink{11},\afflink{19},\afflink{20}},
    Olivier Guyon\textsuperscript{\afflink{21},\afflink{22},\afflink{23},\afflink{24}},
    Sebastiaan Y. Haffert\textsuperscript{\afflink{21}},
    Olivier Herscovici-Schiller\textsuperscript{\afflink{19}},
    \newauthor
    Adrien Hours\textsuperscript{\afflink{16}},
    Roser Juanola-Parramon\textsuperscript{\afflink{25},\afflink{26}},
    Elina Kleisioti\textsuperscript{\afflink{7}},
    Lorenzo König\textsuperscript{\afflink{13}},
    Mariya Krasteva\textsuperscript{\afflink{27}},
    \newauthor
    Iva Laginja\textsuperscript{\afflink{11}},
    Rico Landman\textsuperscript{\afflink{7}},
    Lucie Leboulleux\textsuperscript{\afflink{16}},
    David Mouillet\textsuperscript{\afflink{16}},
    Mamadou N’Diaye\textsuperscript{\afflink{28}},
    \newauthor
    Emiel H. Por\textsuperscript{\afflink{29}},
    Laurent Pueyo\textsuperscript{\afflink{29}},
    Frans Snik\textsuperscript{\afflink{7}},
    Daphne M. Stam\textsuperscript{\afflink{30}},
    Victor Trees\textsuperscript{\afflink{31},\afflink{32}},
    Dirk van Dam\textsuperscript{\afflink{7}},
    \newauthor
    Kyle van Gorkom\textsuperscript{\afflink{21}},
    Maaike van Kooten\textsuperscript{\afflink{33}}
    \newauthor \\%
    Affiliations are listed at the end of the paper
    % Timmy: I checked MNRAS and it seems that this is how they handle very long author lists. See, for example:
    % https://academic.oup.com/mnras/article/498/2/2354/5900562
}
\date{Draft version: \today}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\hwo), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering phenomena. 
%
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
%
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
%
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
%
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
%
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
%
The glint signature at planetary phase angles of $\sim\qtyrange{50}{70}{\degree}$ 
%\sim$110--130$^\circ$ 
would be accessible to the Habitable Worlds Observatory in $\sim\num{14}$ systems, and the rainbow of water clouds at phases of \qtyrange{140}{160}{\degree} would be accessible in $\sim\num{42}$ systems, assuming a \qty{62}{\mas} inner working angle (\IWA{}; $3\,\lambda/D$ for a \qty{6}{\meter} telescope at \qty{600}{\nano\meter}).
%
Increasing the \IWA{} to $2\,\lambda/D$ increases the number of systems with detectable rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA{}.  \end{abstract}

@@

%--------------------------------------------------------------------------------
\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{figures/Fig1IWAPol.png}
    \caption{
        The phase curves of two homogeneous planets with oceans and Earth-like atmospheres in edge-on orbits (the planets thus exhibit phase angles 
        ranging from 0$^\circ$ to 180$^\circ$ along their orbits). 
        The cloudy scenario includes a low-altitude, Earth-like water 
        cloud deck. 
        Below the plot, we are showing the phase angle ranges that cover the
        characteristic features of glories, rainbows, Rayleigh scattering, glint, and forward scattering by cloud particles (note that features partly overlap
        with each other).  
        Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (scattering angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest scattering angles).
    }
    \label{fig:bottplot}
\end{figure*}
%--------------------------------------------------------------------------------

Variations of the flux and in particular the degree of polarization of
light of the parent star that is reflected by an exoplanet as it moves
along its orbit that are due to an ocean glint or optical phenomena from scattering by condensates --- like the rainbow --- could be some of the most distinct signatures of habitability.
%
However, depending on system architectures, some of these features can be obscured by the inner working angle (\IWA{}) of a telescope's coronagraph.

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star, for which a rocky planet (or moon) may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the stellar flux that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and, for example, geological activity. 
%
Establishing the presence of liquid surface water on an exoplanet would therefore provide important context for the interpretation of possible biosignatures.
%
Providing additional context to the habitability of an exoplanet would be the presence of a water cycle, as evidenced by the identification of a glint, i.e.\ specular reflection by a liquid surface, or by a rainbow and other optical phenomena that can be attributed to water clouds, as detailed in the next sections.

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off smooth liquid surfaces. While glint is a familiar phenomenon on Earth, there is only
one other body in the Solar System where it has been observed: 
the Cassini orbiter captured the glint of sunlight reflecting off the
liquid hydrocarbon lakes near Titan's north pole as it was observing
the moon from the nightside \citep{2010GeoRL..37.7104S}.
%
In the total flux, the glint signal is indeed most prominent at small planetary phase (large scattering) angles.
%
The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at \qty{53}{\degree} \citep[i.e., a scattering angle of \qty{127}{\degree}; see, e.g.,][]{2008Icar..195..927W}.
%
This angle therefore varies slightly with wavelength and with sea and atmosphere composition.
%
In general, the glint on a planet will be strongest at red wavelengths where the Rayleigh scattering in the planet's atmosphere and attenuation through the atmosphere are smallest \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also small \citep{2012ApJ...752L...3C}.
%
The strengthened glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
{\color{red}{check the next sentence:}}
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across scattering angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

@@

Rainbows, glories and other optical phenomena from light that is 
scattered by atmospheric condensates provide a wealth of information 
about exoplanets \citep{karalidi2012rainbow, stam2008,2014A&A...566L...1G}.%
\footnote{
    Note that these are not the familiar rainbows that we see when sunlight is scattered by rain droplets, which are huge particles compared to the wavelength of the light: with our eyes, we cannot see the rainbows that are due to sunlight that is scattered by cloud droplets and that are the focus of this paper. 
    However, the presence of these bows (which are also referred to as \enquote{cloudbows}) can be revealed by polarimetric measurements.
}
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, the phase curve behaviour of Venus showed not only that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture but also constrained the droplet size distribution \citep{hansenhovenier1974}. 
The glory feature in Venus's phase curve was later used to rule out some solutions for the unknown UV absorber \citep{petrova2018glory}.

In unpolarized light {\color{Green}{do you mean polarized light?}}, the scattering by cloud particles can dramatically change the shape of a planet's phase curve compared to that of a 
Lambertian reflecting planet.
%
{\color{red}{check the sentence:}}
While the glory and rainbow are technically visible in any phase curve
in the presence of spherical cloud and/or haze particles, their strongly polarized nature coupled with the suppression of signals from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized light (both in the degree of polarization and in full Stokes measurements) \citep{karalidi2011, stam2008, treesandstam2019}.
%
The rainbow angle depends on the refractive index of the scatterers and thus depends on the composition and the wavelength. 
The wavelength dependence itself also depends on the particles' size distribution, and the size distribution affects the angular shape of the rainbow \citep{karalidi2011}
%
Because the rainbow feature is more narrow than the glint, it would 
allow for a more exact determination of the particle composition.
%
Note that the scattering by clouds that consist of solid particles, 
such as ice crystals, will produce other optical features, such as halo's, particularly at small scattering angles \citep{bailey2007, hansentravis1974, karalidi2012rainbow}.
%

@@

%--------------------------------------------------------------------------
\begin{table}
    \centering
    \caption{
        Conversion between $\lambda / D$ and \unit{\mas} for wavelengths $\lambda$ of \qty{600}{\nano\meter} and \qty{1}{\micro\meter}, assuming the telescope's diameter $D$ is $\qty{6}{\meter}$ for the Inner and Outer Working Angles (i.e., the \IWA{} and \OWA{}, respectively). 
        %\todo{We should use consistent wavelengths for this table, for the Bott Plot, and for the M\&M cumulative distribution function. Also update to use the only wavelength we choose.}
    }
    \label{tab:IWA_OWA}
    \begin{tabular}{ c c c c } 
    \toprule
    & $\lambda/D$ & \unit{\mas} (at \qty{600}{\nano\meter}) & \unit{\mas} (at \qty{1}{\micro\meter}) \\
    \midrule
    \midrule
    IWA & 1 & 20.63 & 34.38 \\
    IWA & 2 & 41.25 & 68.75 \\
    IWA & 3 & 61.88 & 103.13 \\
    IWA & 4 & 82.51 & 137.51 \\
    \midrule
    OWA & 32 &  660.05 & 1100.08 \\
    OWA & 64 & 1320.09 & 2200.16 \\
    \bottomrule
    \end{tabular}
\end{table}
%--------------------------------------------------------------------------

\hwo{} is envisioned to observe the stars in the \emph{NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory}.%
\footnote{Available online: \url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list comprises $\sim\num{160}$ stars, the majority of which are Sun-like dwarfs: 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude of $R_c = 31$ and a contrast floor of \num{2.5e-11}. 
%with an occurrence rate of rocky planets in the optimistic habitable zone to be $\eta_{\oplus}$ = 0.24 \citep{decadal}. 
The corresponding adopted habitable zone limits are a semi-major axis of \qtyrange{0.95}{1.67}{\au} for a solar twin, planet sizes between \qtyrange{0.8}{1.4}{} Earth radii, and for non-solar stars, scale as square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}. 
%
The authors of the Decadal Survey determined that for an imaging survey of habitable zones with a \qty{6}{\meter} telescope, these are the nearby stars (maximum distance \qty{25}{\parsec}) around which exo-Earths would be the best observable.

@@

In the non-eccentric case (i.e., a circular orbit with the coronagraphic mask centred on the star), the minimum ($\varphi_\mathrm{min}$) and maximum ($\varphi_\mathrm{max}$) accessible scattering phase angles are symmetric about 90$^\circ$. 
%
We define the angle $\Delta \varphi$ such that: 
\begin{equation}
 \label{eq:Delta_phi}
    \Delta \varphi 
    = \varphi_\mathrm{max} - \varphi_\mathrm{min}
    =  2 \cdot (\varphi_\mathrm{max} - \qty{90}{\degree}) 
    =  2 \cdot (\qty{90}{\degree} - \varphi_\mathrm{min} )  \,.
\end{equation} 
The on-sky projected planet distance, $r_\mathrm{proj}$ (in \unit{\au}) is shown in \cref{fig:scattering-angle} for a non-eccentric case with a 
semi-major orbital axis $a$ (in \unit{\au}), at a distance $d_*$ (in \unit{\parsec}) from the observer. 
%
The scattering phase angle $\varphi$ is given by
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a} \,.
\end{equation}
%
Using small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the angular separation on-sky.
%
We access the minimum scattering angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Using \cref{eq:Delta_phi}, we deduce:
%
\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \varphi}{2}\right) = \frac{\mathrm{IWA} 
    \cdot d_*}{a} \,.
\end{equation}

@@

Most planetary systems will not be viewed as edge-on from Earth.
%
For a system with inclination angle $i$, we can identify the following two regimes (see also \cref{fig:orb-grid}):
\begin{enumerate}
    \item The coronagraph does not obscure any part of the sky-projected orbit: 
    the maximum scattering phase angle coverage depends solely on $i$. This 
    holds for orbits that are relatively face-on; 
    \item The coronagraph does obscure parts of the sky-projected orbit: 
    the scattering phase angle coverage depends only on the \IWA{} and 
    the orbital semi-major axis $a$. 
\end{enumerate}

@@

The NASA Exoplanet Exploration Program’s Mission Star List for \hwo{} 
includes for every star an estimated contrast for an Earth-twin in 
the habitable zone.
%
% The cited paper has all the equations that lead to this equation:
The contrast is given by \citep[see, e.g.,][]{2023A&A...671A.165M}
\begin{equation}
    C(\alpha) 
    = F_p(\alpha) / F_* 
    = A_\mathrm{G} \cdot \phi(\alpha) \cdot \left( R_p / a(\alpha) \right)^2 \,,
\label{eq:contrast}
\end{equation}
where $A_\mathrm{G}$ is the planet's geometric albedo, $R_p$ its radius, $\phi(\alpha)$ is the integral phase function at phase angle $\alpha$
(normalized to \num{1.0} at full phase), and $a$ is the distance between the planet and its host star.
\todo{here, $\alpha$ is used for the phase angle ... and is the 
'integral phase function' the same as the 'scattering phase function'?}
%
The assumptions that went into the contrast estimations for \hwo{}
are that $a$ is constant (the orbits are thus circular), the geometric albedo $A_{\rm G}$ is \num{0.2}, and that the scattering phase function, $\phi(\alpha)$, can be described by a simple Lambertian reflectance phase function, which assumes isotropic reflection by the planetary surface.
%
These calculations give an order-of-magnitude calculation for the 
contrast of Earth-like planets around nearby stars.
%
Therefore, in this work, we will assume that the calculated contrasts 
in the list correspond to the reflected flux at quadrature and normalize the scattering phase functions to this point.
%
We do not aim to accurately calculate the absolute scaling of the contrast, as this depends on many factors and would have required some work for the star list to be redone.
%
